### 2018-04-26 Paper/jaccardbaetica.tex
add introduction

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. They allow us to highlight a part of mechanism of production and distribution of goods between different communities \citep{bevan_mediterranean_2014}.

%This paper explore--- resumir en dos líneas el objeto de estudio y definir que ha sido usado para ello
This paper aims to understand the production dynamics in relation to a specific area within Roman Empire. We want to detect the pattern of olive oil production that link amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production centres. An ecology approach has been used to analyse the dispersion of stamps between amphora workshops.

@@

% Debate
Despite Roman studies have widely discussed in the last centuries, the knowledge about the production processes is still under debate mostly due to the lack of written sources (CAMBIAR PORQUE LO REPITO OCHENTA VECES). The application of different approaches combined with the archaeological evidence has allowed us partly to deal with the complexity of understanding the Roman trade \citep{orengo_seeds_2016,brughmans_roman_2016,coto-sarmiento_maria_bayesian_????}. Specifically, our work pretend to detect economic processes in a microescala focuses on a commercial product from a specific province \citep{isaksen_network_2006}.

%más específico--- olive oil

@@

The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, hundred of amphora workshops were created to support the high demand of Roman Empire. These workshops were located along the river Guadalquivir and its tributary Genil supplying from the riverine connectivity to the Mediterranean and Atlantic \citep{garcia_vargas_enrique_formal_2010}.

The majority of olive oil amphorae produced in this province and shipped thought the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps and \textit{tituli picti} and inked in graffities with different information but there is not a general consensus about the meaning of them \citep{rodriguez_baetican_1998}. Stamps are the most studied in this type of amphorae. There is evidence that stamps were used for almost three centuries. (economía oleica betica). Frequently, stamps were marked mainly in handles but rarely in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criteria (CITAR). Stamps was mostly formed for a code of three letters. There letters can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. 
The meaning of the amphora stamps is still under debate. Some authors suggest that they were identified as the land-owners of the olive groves \citep{rodriguez_economioleicola_1977}. Other authors propose that stamp could be the owners of the making-amphorae workshop (CITAR) or even a production counting system \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps defined somehow the system of working in the workshops.

%Hyphotesis
However, some challenges remain under discussion such as how this province was organized and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow some distribution pattern? Did stamps share the same workshop? Neither written evidences have been found that it could explain the economic role of Baetica province in the Roman organization. On the other hand, archaeological evidence show a highly specialized production with a long activity in a this specific area with apparently few changes \citep{remesal_anforas_2004}.

>>>>

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. In archaeology, they allow us to highlight a part of mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of those factors have an important impact during the Roman Age, when the progressive exploitation of communication networks allowed the a major interaction between communities \citep{orengo_seeds_2016}. As a result, an important mechanism of production control under Roman government was created spreading by different areas attending to the richness of different places.

Despite this issue has been widely discussed over many decades, the understanding about the production processes is still under debate, mostly due to the lack of written sources. The application of different approaches combined with the archaeological evidence has allowed us partly to deal with the complexity of understanding the Roman production \citep{orengo_seeds_2016,brughmans_roman_2016,coto-sarmiento_maria_bayesian_????}.

%This paper explore--- resumir en dos líneas el objeto de estudio y definir que ha sido usado para ello dispersion of culture
This paper aims to understand the production dynamics in relation to a specific area within Roman Empire. Specifically, our work pretend to detect microeconomic processes focused on a commercial product from a specific province \citep{isaksen_network_2006}. We want to detect the pattern of olive oil production that link amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production centres. To do this, an Ecology approach has been used to analyse the dispersion of stamps between amphora workshops.

@@

% Debate

%más específico--- olive oil
%incluir algo que enlace con roman province

\section{The amphora production in the Roman Empire}

@@

The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, hundred of amphora workshops were created to support the high demand of Roman Empire. These workshops were located along the rivers supplying from the riverine connectivity to the Mediterranean and Atlantic \citep{garcia_vargas_enrique_formal_2010}.

The majority of olive oil amphorae produced in this province and shipped thought the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps and \textit{tituli picti} and inked in graffities with different information but there is not a general consensus about the meaning of them \citep{rodriguez_baetican_1998}. Stamps are the most studied in this type of amphorae. There is evidence that stamps were used for almost three centuries. (economía oleica betica). Frequently, they were marked mainly in handles but rarely in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criteria (CITAR). Stamps was mostly formed for a code of three letters. There letters can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

%Many hypothesis involved the meaning of stamps. 
The meaning of the amphora stamps is not clear. On the one hand, stamps are identified as the land owner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could be the owner of the making amphorae workshops or even a production counting system  \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps became in a good proxy to define somehow the system of working in the workshops.

%Hyphotesis
However, some challenges remain under discussion such as how this province was organized and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow some distribution pattern? Did stamps share the same workshop? Neither written sources have been found that it could explain the economic role of Baetica province in the Roman organization. On the other hand, archaeological evidence show a highly specialized production with a long activity in a this specific area with apparently few changes \citep{remesal_anforas_2004}.
